---
title: "Analysis of Weather Event Types on US Population Health and Economics"
author: "Charlie Astrada"
date: "10/29/2016"
output: html_document
---

## Synopsis

The purpose of this report is to explore the US National Oceanic and Atmospheric Administration's (NOAA) storm database and answer the following questions: 

* Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

The report concludes that: 

* Tornadoes are the most harmful weather event type to population health in the US  
* Floods are the most economically damaging weather event type in the US


## Data Processing

The libraries below are the ones we will need for our analysis work. 

```{r load libraries, message=FALSE}
library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)
```


Data is loaded into a data frame and only the variables that we require are selected in order to speed up working with the data frame. 
```{r load data, cache=TRUE}
# load data into dataframe
storm.data <- read.csv("repdata%2Fdata%2FStormData.csv.bz2", na.strings = c("", " ", "NA"), stringsAsFactors = F)

# select only those variables required for the analysis
storm.data <- storm.data %>% 
        select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

The `EVTYPE` variable contains the relevant weather event types. There are many misspellings and similarly named values. We will want to perform some cleanup and grouping to tidy things. 

```{r explore data}
head(unique(storm.data$EVTYPE), n = 100)
```

The code below groups and cleans up the `EVTYPE` data.
```{r group data, cache=TRUE}
# make all values lower case and trim white space to simplify regexps
storm.data$EVTYPE <- str_to_lower(storm.data$EVTYPE)
storm.data$EVTYPE <- str_trim(storm.data$EVTYPE, side = "both")

# Group EVTYPEs based on best guesstimate relevance
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(wnd|wns|wind).*", replacement = "wind")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(snow|icy|ice|winter weather|wintery|freeze|sleet|frost).*", replacement = "snow_ice_frost")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(blizzard).*", replacement = "blizzard")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(winter storm|freezing drizzle|winter mix|wintry).*", replacement = "winter_storm")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(hail).*", replacement = "hail")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(tornado|torndao).*", replacement = "tornado")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(hurricane|floyd).*", replacement = "hurricane")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(fldg|fld|flood|floood).*", replacement = "flooding")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(wall|funnel).*", replacement = "wall_cloud/funnel_cloud")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(record high|record warm|very warm|high temp).*", replacement = "high_temperature")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(hot|warm|heat|unusual warmth).*", replacement = "high_temperature")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(record low|low temp|cold|cool).*", replacement = "low_temperature")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(dry|driest|drought).*", replacement = "dry_weather")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(microburst).*", replacement = "dry_microburst")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(thunderstorm|tstm|storm|gustnado).*", replacement = "thunderstorm")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(rain|precipitation|precipatation|precip|heavy mix|shower|wet).*", replacement = "wet_weather")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(erosion|erosin).*", replacement = "erosion")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(fog|vog).*", replacement = "fog")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(smoke).*", replacement = "smoke")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(fire).*", replacement = "fire")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(landslide|landslump).*", replacement = "landslides")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(tide).*", replacement = "tide")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(surf).*", replacement = "surf")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(wave).*", replacement = "wave")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(dam).*", replacement = "dam_failure")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(water spout|waterspout|wayterspout).*", replacement = "waterspout")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(lightning|ligntning|lighting).*", replacement = "lightning")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(dust).*", replacement = "dust")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(swells|sea).*", replacement = "rough_sea")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(temperature record).*", replacement = "record_temperature")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(volcanic).*", replacement = "volcanic_activity")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(marine mishap).*", replacement = "marine_accident")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(mud).*", replacement = "mudslide")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(urban|stream).*", replacement = "stream")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(avalance|avalanche).*", replacement = "avalanche")
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(seiche).*", replacement = "stream")
# group nonsensical EVTYPEs into other
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = ".*(summary|southeast|monthly temperature|red flag criteria|apache|excessive).*", replacement = "other") 
# Group unknowns into other
storm.data$EVTYPE <- str_replace(storm.data$EVTYPE, 
                                 pattern = "\\?", replacement = "other") 
```

The `PROPDMGEXP` and `CROPDMGEXP` are coded exponents corresponding to hundreds (H), thousands (T), millions (M), and billions (B). These exponents must be multiplied by the `PROPDMG` and `CROPDMG` values respectively to come up with single totals. Note that `PROPDMGEXP` and `CROPDMGEXP` contain superfluous codes, which we ignore. 

```{r calculate  totals}
# make all codes upper case for simplicity
storm.data$PROPDMGEXP <- str_to_upper(storm.data$PROPDMGEXP)
storm.data$CROPDMGEXP <- str_to_upper(storm.data$CROPDMGEXP)

# Create a new variable containing the property damage multipled by the exponent for simplicity
storm.data <- storm.data %>%
        mutate(PROP_DMG_TOTAL = if_else(PROPDMGEXP == "H",
                                        PROPDMG * 1e+2,
                                        if_else(PROPDMGEXP == "K", 
                                                PROPDMG * 1e+3, 
                                                if_else(PROPDMGEXP == "M",
                                                        PROPDMG * 1e+6,
                                                        if_else(PROPDMGEXP == "B",
                                                                 PROPDMG * 1e+9,
                                                                 PROPDMG)))))

# Create a new variable containing the crop damage multipled by the exponent for simplicity
storm.data <- storm.data %>%
        mutate(CROP_DMG_TOTAL = if_else(CROPDMGEXP == "H",
                                        CROPDMG * 1e+2,
                                        if_else(CROPDMGEXP == "K", 
                                                CROPDMG * 1e+3, 
                                                if_else(CROPDMGEXP == "M",
                                                        CROPDMG * 1e+6,
                                                        if_else(CROPDMGEXP == "B",
                                                                CROPDMG * 1e+9,
                                                                CROPDMG)))))


```

In the plot below we take a top 5 ranking by total injuries/fatalities using an overlaid bar chart. We can see that tornadoes are by far the biggest hazard to population health. 
```{r plot fatalities injuries}

# Summarise dataframe with injuries, fatalities, and total of two
inj_fatalities <- storm.data %>%
        group_by(EVTYPE) %>% 
        summarise(total_injuries_fatalities = sum(FATALITIES,INJURIES, na.rm = T),
                  injuries = sum(INJURIES, na.rm = T),
                  fatalities = sum(FATALITIES, na.rm = T))

# create a top 5 rank by total injuries/fatalities
top_5_by_inj_fatalities <- inj_fatalities %>% 
        top_n(n = 5, wt = total_injuries_fatalities) 


# we need to adjust data frame to plot both fatalities and injuries
top_5_by_inj_fatalities <- gather(top_5_by_inj_fatalities, key = event_type,
                                   value = total,
                                   fatalities, injuries)

p.health <- ggplot(top_5_by_inj_fatalities, aes(EVTYPE, total, fill = event_type))
p.health + geom_bar(stat = "identity") + 
        labs(title = "Top 5 Events by Fatalities and Injuries", x = "Weather Event", y = "Total Persons") +
        scale_fill_discrete(name = "Weather Event")


```


In the plot below, we take a top 5 ranking by property and crop damage. We see that flooding is by far the largest economic hazard. 

```{r plot property and crop damage}

# Summarise dataframe with property and crop damage total
economic_damage <- storm.data %>%
        group_by(EVTYPE) %>% 
        summarise(total_economic_damage = sum(PROP_DMG_TOTAL, CROP_DMG_TOTAL, na.rm = T))

# create a top 5 rank by total economic damage
top_5_by_economic_damage <- economic_damage %>% 
        top_n(n = 5, wt = total_economic_damage) 


p.health <- ggplot(top_5_by_economic_damage, aes(EVTYPE, total_economic_damage, fill = EVTYPE))
p.health + geom_bar(stat = "identity", show.legend = F) + 
        labs(title = "Top 5 Events by Economic Damage", x = "Weather Event", y = "Total US Dollars")

```

## Results

All totals are rounded where appropriate. 

### Population Health
Tornadoes are the most harmful weather event type to US population health with a total of 97043 injuries or fatalities. 

The remaining 4, ordered by total fatalities or injuries, are: 

2. wind:                12949  
3. high temperatures:   12421  
4. flooding:            10234  
5. lightning:           6048


### Economic Health

Flooding is the most economically damaging weather event type with a total of 180B USD in total damage. 

The remaining 4, ordered by most economic damage, are: 

2. hurricanes:          90B USD  
3. thunderstorms:       64B USD  
4. tornadoes:           57B USD  
5. wind:                20B USD  
